NanoMaLy

### 2015-11-15 paper/pldi16-submission/semantics.tex
evaluation
confusing terminology

% \begin{itemize}
% \item how do we run ill-typed programs?
% \item for a lang like ocaml, dynamic semantics are independent of static
%   semantics, just lambda calculus. so no problem to run ill-typed
%   program
% \item but what about functions? what type of arguments should we pass? consider
%
% \begin{lstlisting}
% let f x = 
%   let y = 1 + x in
%     1. +. y
% \end{lstlisting}
%
% does \texttt{f} take an int, float, string? int and float are both
% somewhat plausible, but string or anything else is ``clearly'' bogus. so
% we cannot provide \emph{completely arbitrary} inputs to
% \texttt{f}. Instead, we call \texttt{f} with a \emph{hole}, written
% \ehole{}, which is a placeholder for a value whose type we have not
% yet determined. As we execute the program, we instantiate holes with
% concrete values as demanded by the primitive operations in the
% program. For example, the hole we pass to f will be instantiated to an
% int when we reach the \lstinline{1 + x} term. Thus, y will be an int as
% well, and the program will get stuck at \lstinline{1. +. y}. \ES{this
%   reads more like overview text..}
%
% % \item values are tagged with their types, just like ``untyped'' langs
% % \item special ``hole'' value whose type is not yet known, used for function args
% % \item on-the-fly unification to determine ``correct'' type for holes
% \end{itemize}
%
Next, we formalize our search for witnesses to type-errors.
%
We present the syntax and operational semantics of \lang -- a simple
lambda calculus with integers and booleans, extended with our notion of
holes -- as well as our search algorithm.
%
We prove that our system \emph{soundly} finds witnesses, \ie if we find
a witness then there is no possible typing for the input program.
%
\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with numbers and booleans.
%
As we are specifically interested in
programs that \emph{do} go wrong, we include an explicit \stuck state in
our syntax.
%
\paragraph{Holes}
\label{sec:holes}
The main novelty in our system is the notion of a ``hole'', written
\ehole{}, which represents an uninstantiated value.
%
Importantly, we do not even know what type the value should have.
%
Holes may also appear in types, where they may be thought of as type
variables that we will not generalize over.
%
\subsection{Semantics}
\label{sec:semantics}
\input{operational.tex}
%
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
We write \stepi{j}{e}{\su}{e'}{\su'} if there exist $e_1,\ldots,e_j$ such that
$e$ is $e_1$, $e'$ is $e_j$ and $\forall i,j, 1 \leq i < j$, we have
\step{e_i}{\su_i}{e_{i+1}}{\su_{i+1}}.
%
We write \steps{e}{\su}{e'}{\su'} if there exists some (finite) $j$ such that
$\stepi{j}{e}{\su}{e'}{\su'}$.
%
The evaluation relation is parameterized by a pair of \forcesym and
\gensym functions that handle runtime type-checking and instantiation of
holes respectively.
%
The relation must also maintain a substitution \su, mapping holes to
concrete values, so that we do not instantiate the same hole with
different values in different contexts, and so that we can report a
concrete witness to any discovered type errors.

@@

  We will omit the subscript when we wish to refer to the final partial
  type.
\end{defn}
%
The \typeof{} function is an approximation of the type of an expression.
\begin{defn}
\label{def:typeof}
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun{\thole{i}}{\typeof{e}}, & \quad \text{$i$ is fresh} \\
    \typeof{e} & \defeq & \thole{i}, & \quad \text{$i$ is fresh} \\
  \end{array}
  \]
\end{defn}
%
We also define a compatibility relation between types.
%
\begin{defn}
\label{defn:type-compat}
  A type $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t}, if
  $\exists \su.\ t = \subst{\su}{s} \lor s = \subst{\su}{t}$.
\end{defn}
%
Given these two definitions, we show that each evaluation step
refines the partial type of $f$.
%
\begin{lem}
\label{lem:refine-partial}
  For all $k$, \tcompat{\ptype{k}{f}}{\ptype{k+1}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules. 
  %
  Note that all rules are type-preserving with the exception of when
  \forcesym is called on a hole, which case we instantiate the hole with
  a concrete value.
  %
  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
\end{proof}
%
Furthermore, only a call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
  For all $k$, if $\ptype{k}{f} \neq \ptype{k+1}{f}$, then \forcesym must
  have been called at step $k+1$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules. 
  % 
  If $\ptype{k}{f} \neq \ptype{k+1}{f}$ then one of the holes in $f$'s
  argument must have been instantiated with a concrete value at step
  $k+1$.
  %
  An examination of the rules shows that only place this happens is 
  in the first case of \forcesym.
\end{proof}
%
Finally, we show that any value that is not compatible with the $k$th
partial type of $f$ will cause $f$ to get stuck in \emph{at most} $k$
steps.
%
\begin{lem}
\label{lem:k-stuck}
  For any \hastype{v}{t} that is not compatible with
  the input of \ptype{k}{f}, \stepi{k}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\begin{proof}
  By induction on $k$.

@@

      -- (1) find the correct number of arguments
      loop :: [$v$] -> ([$v$], [($v$, $\su$)])
      loop args = case eval (build args, []) of 
        ($\efun{x}{e}$, _) : _ -> loop (args `snoc` $\ehole{}$)
        results      -> (args, results)  
  \end{mcode}
  \caption{Implementing our counter-example search in terms of the
    operational semantics. 
    \ES{1 and 2 are backwards, ugh...}
    \ES{should address case where output types of successive runs dont match}
  }
  \label{fig:search-algo}
\end{figure*}
%
Figure~\ref{fig:search-algo} summarizes the overall implementation of
our search for witnesses, which takes as input a sequences of binders
and returns either @Safe@ if no witness could be found, or |Unsafe $e$|
where $e$ is a term that causes the input program to get stuck.

>>>>

\section{Searching for Type-Error Witnesses}
\label{sec:searching-witness}
% \begin{itemize}
% \item how do we run ill-typed programs?
% \item for a lang like ocaml, dynamic semantics are independent of static
%   semantics, just lambda calculus. so no problem to run ill-typed
%   program
% \item but what about functions? what type of arguments should we pass? consider
%
% \begin{lstlisting}
% let f x = 
%   let y = 1 + x in
%     1. +. y
% \end{lstlisting}
%
% does \texttt{f} take an int, float, string? int and float are both
% somewhat plausible, but string or anything else is ``clearly'' bogus. so
% we cannot provide \emph{completely arbitrary} inputs to
% \texttt{f}. Instead, we call \texttt{f} with a \emph{hole}, written
% \ehole{}, which is a placeholder for a value whose type we have not
% yet determined. As we execute the program, we instantiate holes with
% concrete values as demanded by the primitive operations in the
% program. For example, the hole we pass to f will be instantiated to an
% int when we reach the \lstinline{1 + x} term. Thus, y will be an int as
% well, and the program will get stuck at \lstinline{1. +. y}. \ES{this
%   reads more like overview text..}
%
% % \item values are tagged with their types, just like ``untyped'' langs
% % \item special ``hole'' value whose type is not yet known, used for function args
% % \item on-the-fly unification to determine ``correct'' type for holes
% \end{itemize}
%
Next, we formalize our search for witnesses to type-errors.
%
We present the syntax and operational semantics of \lang -- a simple
lambda calculus with integers and booleans, extended with our notion of
holes -- as well as our search algorithm.
%
We prove that our system \emph{soundly} finds witnesses, \ie if we find
a witness then there is no possible typing for the input program.
%
\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with numbers and booleans.
%
As we are specifically interested in
programs that \emph{do} go wrong, we include an explicit \stuck state in
our syntax.
%
\paragraph{Holes}
\label{sec:holes}
The main novelty in our system is the notion of a ``hole'', written
\ehole{}, which represents an uninstantiated value.
%
Importantly, we do not even know what type the value should have.
%
Holes may also appear in types, where they may be thought of as type
variables that we will not generalize over.
%
\subsection{Semantics}
\label{sec:semantics}
\input{operational.tex}
%
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
We write \stepi{j}{e}{\su}{e'}{\su'} if there exist $e_1,\ldots,e_j$ such that
$e$ is $e_1$, $e'$ is $e_j$ and $\forall i,j, 1 \leq i < j$, we have
\step{e_i}{\su_i}{e_{i+1}}{\su_{i+1}}.
%
We write \steps{e}{\su}{e'}{\su'} if there exists some (finite) $j$ such that
$\stepi{j}{e}{\su}{e'}{\su'}$.
%
The evaluation relation is parameterized by a pair of \forcesym and
\gensym functions that handle runtime type-checking and instantiation of
holes respectively.
%
The relation must also maintain a substitution \su, mapping holes to
concrete values, so that we do not instantiate the same hole with
different values in different contexts, and so that we can report a
concrete witness to any discovered type errors.

@@

  We will omit the subscript when we wish to refer to the final partial
  type.
\end{defn}
%
The \typeof{} function is an approximation of the type of an expression.
\begin{defn}
\label{def:typeof}
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun{\thole{i}}{\typeof{e}}, & \quad \text{$i$ is fresh} \\
    \typeof{e} & \defeq & \thole{i}, & \quad \text{$i$ is fresh} \\
  \end{array}
  \]
\end{defn}
%
We also define a compatibility relation between types.
%
\begin{defn}
\label{defn:type-compat}
  A type $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t}, if
  $\exists \su.\ t = \subst{\su}{s} \lor s = \subst{\su}{t}$.
\end{defn}
%
Given these two definitions, we show that each evaluation step
refines the partial type of $f$.
%
\begin{lem}
\label{lem:refine-partial}
  For all $k$, \tcompat{\ptype{k}{f}}{\ptype{k+1}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules. 
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym is called on a hole, which case we instantiate the hole with
  a concrete value.
  %
  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
\end{proof}
%
Furthermore, only a call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
  For all $k$, if $\ptype{k}{f} \neq \ptype{k+1}{f}$, then \forcesym must
  have been called at step $k+1$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules. 
  % 
  If $\ptype{k}{f} \neq \ptype{k+1}{f}$ then one of the holes in $f$'s
  argument must have been instantiated with a concrete value at step
  $k+1$.
  %
  An examination of the rules shows that only place this happens is 
  in the first case of \forcesym.
\end{proof}
%
Finally, we show that any value that is not compatible with the $k$th
partial type of $f$ will cause $f$ to get stuck in \emph{at most} $k$
steps.
%
\begin{lem}
\label{lem:k-stuck}
  For any \hastype{v}{t} that is not compatible with
  the input of \ptype{k}{f}, \stepi{k}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\begin{proof}
  By induction on $k$.

@@

      -- (1) find the correct number of arguments
      loop :: [$v$] -> ([$v$], [($v$, $\su$)])
      loop args = case eval (build args, []) of 
        ($\efun{x}{e}$, _) : _ -> loop (args `snoc` $\ehole{}$)
        results      -> (args, results)  
  \end{mcode}
  \caption{Implementing our counter-example search in terms of the
    operational semantics. 
    \ES{1 and 2 are backwards, ugh...}
    \ES{should address case where output types of successive runs dont match}
  }
  \label{fig:search-algo}
\end{figure*}
%
Figure~\ref{fig:search-algo} summarizes the overall implementation of
our search for witnesses, which takes as input a sequences of binders
and returns either @Safe@ if no witness could be found, or |Unsafe $e$|
where $e$ is a term that causes the input program to get stuck.
%
The search is split into two phases: (1) we supply an increasing number
of arguments until we find a saturated application, and (2) we search
through the list of results from the non-deterministic evaluator for a
witness.
